---
title: "Environmental and Demographic Analysis of Pittsburgh, PA"
subtitle: "Proposal"
author: 
  - name: "Uchihas"
    affiliations:
      - name: "School of Information, University of Arizona"
description: "This project investigates the impact of NO2 concentration on hospital admissions and the relationship between floodplain extent and housing density in Pittsburgh, PA, using multiple datasets."
format:
  html:
    code-tools: true
    code-overflow: wrap
    code-line-numbers: true
    embed-resources: true
editor: visual
code-annotations: hover
execute:
  warning: false
  message: false
---

```{r}
#| label: load-pkgs
#| message: false
library(tidyverse)
```

## Dataset

```{r}
#| label: load-dataset
#| message: false
# Load the datasets
benmap <- read_csv("data/PitPA_BenMap.csv")
bg_pop <- read_csv("data/PitPA_BG_Pop.csv")

# Display the dimensions of each dataset
list(
  BenMap_Dimensions = dim(benmap),
  BG_Pop_Dimensions = dim(bg_pop),
)
glimpse(benmap)
glimpse(bg_pop)
```

**Description**:

-   **PitPA_BenMap**: This dataset, sourced from the EPA’s BenMAP program, includes 1089 observations across 15 variables, focusing on air quality (e.g., NO2 levels) and health outcomes such as hospital admissions for respiratory conditions.

-   **PitPA_BG_Pop**: Contains demographic data at the block group level, with 1089 observations and 13 variables, including population counts and housing information. Sourced from the American Community Survey.

    -   **Reason for Choosing**:

        -   These datasets offer a comprehensive view of the environmental and demographic factors influencing Pittsburgh's communities. They enable a dual focus on health impacts of air pollution and the relationship between health risk and population density.

## Questions

The two questions you want to answer.

#\| label: analysis-plan #\| message: false

-   **Research Question 1**: How does the concentration of NO2 & SO2 impact the rate of hospital admissions for respiratory conditions across different block groups in Pittsburgh, PA?

    -   **Importance**: This question is crucial for understanding the direct impact of air pollution on community health, helping to inform public health interventions.

    -   **Variables**:

        -   **NO2 concentration** (Quantitative)

        -   **Rate of hospital admissions for respiratory conditions** (Quantitative)

        -   **Block group identifier** (Categorical)

-   **Research Question 2**: How does the extent of land in floodplain zones affect housing density across different block groups in Pittsburgh, PA?

    -   **Importance**: To analyze whether certain age groups or racial demographics are disproportionately affected.

    -   **Variables**:

        -   **Age Groups: under_1, under_13, over_70, and their percentages.** (Quantitative)

        -   **Health Metrics: E.g., NO2_Asthma_Exacerbation_V, O3_Mortality_V, PM25_Mortality_V** (Quantitative)

        -   **Racial Demographics: NonWhite, NonWt_Pct** (Categorical)

------------------------------------------------------------------------

## Analysis plan

-   A plan for answering each of the questions including the variables involved, variables to be created (if any), external data to be merged in (if any).

    #\| label: analysis-plan #\| message: false

    -   **Analysis Plan for Research Question 1**:

        -   **Data Preparation**:

            -   Clean and filter the `benmap` dataset to retain relevant columns: `NO2 concentration`, `hospital_admissions`, and `bgrp`.

            -   Merge with `bg_pop` to incorporate demographic controls if necessary.

        -   **Visualization**:

            -   Create a scatter plot of NO2 concentration against hospital admissions using `ggplot2`.

            -   Use `geom_smooth()` to add a trend line and assess the strength of the relationship.

            -   Use `leaflet` or `ggplot2` with spatial data to create a heatmap overlaying NO2 levels and hospital admission rates.

        -   **Modeling**:

            -   Use linear regression to model the relationship between NO2 concentration and hospital admissions.

            -   Include demographic variables (e.g., population density) from `bg_pop` as controls to adjust for confounding factors.

    -   **Analysis Plan for Research Question 2**:

        -   **Data Preparation**:

            -   Join the datasets on Block_Group and bgrp.

            -   Analyze correlations between demographic variables (e.g., age, race) and health impacts to identify disproportionate effects.

        -   **Visualization**:

            -   Create a scatter plot of the percentage of Age Groups against housing density.

            -   Use `geom_point()` and `geom_smooth()` to observe trends and relationships.

            -   Create a bar chart or box plot to compare housing densities across different ranges of Age Groups/Racial demographics.

        -   **Modeling**:

            -   Use linear regression to analyze the impact of Age group percentage on housing density.

            -   Consider using interaction terms to explore if the relationship differs across urban and suburban areas.

                ### Next Steps

                1.  **Data Cleaning**: Perform initial cleaning to handle missing values, ensure consistent column names, and validate data types.

                2.  **Data Exploration**: Conduct exploratory data analysis (EDA) to understand distributions and potential outliers in the key variables.

                3.  **Merge Datasets**: Use `left_join()` or `inner_join()` in `dplyr` to merge `benmap` and `bg_pop` datasets based on `bgrp` (block group identifier).

                4.  **Visualization and Analysis**: Generate the planned visualizations and build statistical models to test hypotheses.

                5.  **Interpretation**: Interpret the results in the context of urban planning and public health for Pittsburgh, PA.

    ------------------------------------------------------------------------
